Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-05-18 doc/temp-sentiment.tex
Updates from Overleaf

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
% \usepackage[
%   style=numeric,
%   citestyle=numeric-comp,
%   backend=biber,
%   doi=true,
%   natbib=true,
%   sorting=none
% ]{biblatex}
\usepackage{scalerel,xparse}

@@

\author[1, *]{Matthew Cooper}
\author[2]{Jeremiah Osborne-Gowey}
\author[3]{Zheng Liu}
\author[4]{Jie Liu}
\author[5]{Portia Adade Williams}
\author[6]{Aaron Schwartz}
\author[7]{Patrick Baylis}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{University of Colorado Boulder}
\affil[3]{Department of Geographical Sciences, University of Maryland College Park}
\affil[4]{School of Business, East China University of Science and Technology}
\affil[5]{University of Cape Town}
\affil[6]{University of Colorado Boulder}
\affil[7]{University of British Columbia}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in 250 million geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effects is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being with the impacts projected to worsen in the coming years \cite{pachauri2014climate}.  A large body of research has assessed the ongoing impacts on climate change on outcomes like agriculture, economic growth, and physical health (NEED CITATIONS). Academic and medical researchers have highlighted the relative paucity of empirical studies into climate impacts on mental health at multiple scales and recently called for more \cite{Berry2018Apr, hayes_climate_2018}.  Researchers have begun to answer these calls, finding associations between deteriorating environmental conditions and a sense of "ecological grief" from their experienced loss \cite{Cunsolo2018Apr}, disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, as well as a variety of mental health impacts associated with warmer temperatures \cite{baylis_weather_2018, Mullins2019Dec} (PALINKAS and WONG 2020).  MAYBE A SENTENCE HERE HIGHLIGHTING THE GAP IN KNOWLEDGE ABOUT DISPARITIES IN MENTAL HEALTH IMPACTS, VULNERABLE POPULATIONS, issues of scale for signal to noise, etc.

%Heat and mental health
%https://www.pnas.org/content/115/43/10953
Higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other work demonstrating similar relationships in other locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}.  Higher temperatures have also been correlated with increases in hospitalization events related to both general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb, Gupta1992Jun} as well as increased mortality among people with mental health disorders \cite{Hansen2008Oct}.

Temperature has been hypothesized to impact mental health through a number of pathways.  Work on biological mechanisms emphasize that there may be mental health effects from maintaining stable body temperature in high heat \cite{Lohmus2018Jul}.

Additionally, recent studies have found that nighttime temperatures affects sleep quality \cite{Obradovich2017May, Mullins2019Dec}.

% The pathways
Higher temperature associated with climate change could influence mental health both directly and indirectly by exposing people to trauma (direct) (CITATIONS) or by affecting (1) physical health and (2)community well-being (indirect) \cite{BerryETAL2010}. Direct pathways can include extreme heat events or increasing temperatures associated with loss of workplace productivity from decreases in occupational health or increases in workplace injury \cite{kjellstrom_heat_2016, KjellstromETAL2009, kjellstrom_impact_2016} which can result in loss of income, all of which can exacerbate deteriorating mental health and lead to increased aggression, hospital admissions and suicide rates \cite{BerryETAL2010,CheatwoodDerral1995,CohnETAL2004,BouchamaETAL2007}. Heat exposures in working environments could also reduce people's capacity to deal with physical and mental task, and increase the risk of accidents, because of the excessive core body temperature and dehydration \cite{KjellstromETAL2009}. The loss of work capacity would in turn result in loss of income, while the low income could also cause mental health problems \cite{Katz1997}.  As for the indirect pathway, first, because of the reciprocal causation relationship between physical health problems and mental health problems \cite{MillerETAL2009,PrinceETAL2007} heat exposures associated with climate change along with other climate events and indirect health risks threatening physical health will directly influence mental health \cite{Berry2007,WHO2007}. Second, disordered temperature associated with climate change may destroy the economic in agricultural-production-dependent communities. For example, extreme heat reduces the work capacity of laborers in farm fields \cite{KjellstromETAL2009}, which further destroys agricultural-supported industries in local area\cite{BerryETAL2008}. The following economic pressures would undermine social capital and then lead to mental health problems.

Recent research conducted on the effects of higher temperatures on mental health conducted at large, multi-city scales has found surprisingly uniform vulnerability.  For example, focusing on suicides, Burke et al. found ``no significant difference in suicide response to temperature between rich and poor municipalities or counties" \cite{Burke2018Aug}.  Similarly, Mullins et al. find no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status \cite{Mullins2019Dec}.  One major challenge for these studies is that they rely on data from public health agencies that is aggregated to the municipality or city level.  Thus, they must rely on between-county metrics of vulnerability and cannot take neighborhood effects into account.

%We would expect the poor to be more vulnerable, but this is not what we find
The poor and disadvantaged are much more exposed to undesirable temperatures.  They are more likely to work outside rather than indoors; they are more likely to rely on public transportation, bicycle, or walk instead of commute in their own air-conditioned car; and they are more likely to live in housing that is less insulated with poorer quality air-conditioning.  Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, and these heterogeneities in vulnerability might be most pronounced at the neighborhood scale.  Indeed this is what we find for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar}.

While public health data on mental health outcomes is not available at the neighborhood scale, data from twitter can provide an important indicator of mental health at extremely precise spatial and temporal scales.  Many studies have used the mood expressed in tweets as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}.  Such studies have used twitter data to identify the onset of Post Traumatic Stress Disorder in individuals, even before their formal diagnosis \cite{Reece2017Oct}, while similar methods with facebook data have been used to predict the onset of depression \cite{Eichstaedt2018Oct}.  Additionally, in London, day-to-day changes in mental health indicators derived from tweets have been associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}, while across the US more positive tweets are associated with a variety of metrics of human well being \cite{Mitchell2013May}.  Finally, the mood expressed in tweets has been strongly associated with local weather \cite{baylis_weather_2018}, and has been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}.  Thus, we used a common sentiment metric to determine if neighborhood characteristics affect the vulnerability of mental health to higher temperatures.

@@

\subsection*{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  Controlling for a variety of fixed effects, we found that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree Celsius (a dry bulb temperature typically around 12\textdegree C/54\textdegree F).  The greatest declines in sentiment are observed between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We ran models using multiple sentiment indicators, and found similar results no matter which sentiment metric was used (See Supplement).

@@

We then explored how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment (See Fig \ref{fig:sub1}).  We found that neighborhood income strongly moderates the relationship between temperature and sentiment.  As temperatures increase to moderate heat of 20\textdegree C WBGT (29\textdegree C/84\textdegree F), sentiment decreases for most neighborhoods, but increases in neighborhoods in the 95th income percentile.  These rich do not see decreases in sentiment until temperatures are above 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point sentiment decreases evenly across income ranges, still leaving a large gap in expressed sentiment from low to high-income neighborhoods.

We examined patterns of temperature and sentiment for neighborhoods that had a majority population of four broad racial categories - non-Hispanic black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, and Asian-American and Pacific Islander.  While people in neighborhoods of all racial compositions were affected by heat, people in majority black neighborhoods were much more affected than others  (See Fig \ref{fig:sub2}).  Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decreases four-times as much as people in other neighborhoods.  Additionally, at levels of moderate heat from 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap closes with high temperatures.

Finally, we modeled the combined effects of race and income in moderating the impacts of heat on mental health, and we found separate effects for these two variables (See Supplement).  In other words, neither race or income alone account for all the heterogeneity in vulnerability, and neighborhoods that are poor and black are more affected by than than neighborhoods that are just poor or just black.

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-income.png}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-race_q.png}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood income (a) and race (b).}
\label{fig:test}
\end{figure}

@@

Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While a variety of related metrics can be used to quantify sentiment, we chose to use the VADER metric that was specifically designed for microblogs like twitter \cite{hutto2014vader}.  To give context to this novel metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to other events that have large impacts on sentiment (See Fig. \ref{fig:compare}).  Firstly, we examine the change in average sentiment from Saturday to Monday, the highest and lowest days for twitter sentiment as well as other mental health metrics, such as suicides.  We also examine the decrease in sentiment associated with the two most expensive hurricanes in the last decade in the United States.  For the hurricanes, we compared the mean sentiment of counties affected by Hurricanes Harvey and Sandy on the day of landfall to the mean sentiment a week before.  These hurricanes are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

We found that the impacts of heatwaves on sentiment in rich and majority white neighborhoods were less than the average weekly change in sentiment from Saturday to Monday, but the effects of heat waves on majority black or poor neighborhoods were much greater than the average weekly change in sentiment.  Additionally, the impacts of heat waves more vulnerable neighborhoods were close in degree to the impacts of both hurricanes.

{\color{red} What other disasters/comparison events should we look at?}

@@

To further explore our findings, we examined inequalities in the impacts of higher temperatures on sentiment by Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) with over 1 million people (hereafter: cities).  We examined the change in size of the gap in sentiment scores between both poor and rich neighborhoods, as well as black and non-black neighborhoods, as temperatures change from optimum temperatures of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F).

We found increasing inequalities in sentiment across income groups for 68.8\% of cities, with 35.6\% of these having a statistically significant effect (15 cities).  Cities with a significantly unequal effect were found in the south, Midwest, northeast, northwest, and southwest, although they were most common in the mid-Atlantic region.  Additionally, many cities in the southwest had an inverted effect, where higher temperatures actually narrowed the gap in sentiment between rich and poor neighborhoods, including Denver and Oklahoma City, which had a statistically significant inverted effect.

Because we found a much stronger effect for the effects of temperature on sentiment in majority black neighborhoods, we also examined increases in sentiment gaps between black and non-black neighborhoods in cities with a black population greater than 5\% of the total population.  We found increasing inequalities in sentiment for more black neighborhoods in 64.7\% of cities, with a significant effect in 21.2\% of those (7 cities).  Additionally, three sunbelt cities had a significant inverted effect.  The patterns of inequalities in the impacts of heat on sentiment by race were similar to inequalities by income: cities with large and significant inequalities were found the mid-Atlantic and Midwest, as well as central Florida.

\begin{figure}[H]
\centering
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_income.png}
  \caption{}
  \label{fig:map1}
\end{subfigure}
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_black.png}
  \caption{}
  \label{fig:map2}
\end{subfigure}
\caption{Inequalities in impacts of temperature on sentiment by CSAs and MSAs with over 1 million people.  The value shown is the predicted change in the size of the gap in VADER score as temperatures increase from 5WBGT to 30WBGT between the 5th and 95 percentile for income (a) and the 95th and 5th percentile for percent black (b).  CSAs with a statistically significant effect ($p < 0.05$) are outlined in black.  For racial inequalities (b), only MSAs and CSAs where black people make up $> 5\%$ of the population are included.}
\label{fig:test}
\end{figure}

\section*{Discussion}
% Benefits of twitter data/overview of findings
We found that twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution.  We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between heat and worsened sentiment.   Moreover, we were able to show clear heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

% Sentiment is a fuzzy indicator but we have BIG DATA
However, one limitation of data from twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of nearly 250 million tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

% Compare weekly changes in sentiment to weekly changes in suicides
We found that the change from optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to a heatwave of 25\textdegree C WBGT (36\textdegree C/97\textdegree F) is associated with an overall decrease in the VADER sentiment score of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372), a weekly change associated with a increase in suicides of 23\% over the course of the study period \cite{CDC2021}.  Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment and suicides at some time scales.  Moreover, both suicides and sentiment are affected by higher temperatures \cite{baylis_weather_2018, Burke2018Aug}.  Thus, while twitter sentiment is the only mental health indicator available at fine enough spatial and temporal scales to examine neighborhood effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations, although this analysis cannot speak to those outcomes specifically.

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{breakcites}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
% \usepackage[
%   style=numeric,
%   citestyle=numeric-comp,
%   backend=biber,
%   doi=true,
%   natbib=true,
%   sorting=none
% ]{biblatex}
\usepackage{scalerel,xparse}

@@

\author[1,*]{Matthew Cooper}
\author[2]{Jeremiah Osborne-Gowey}
\author[3]{Zheng Liu}
\author[4]{Jie Liu}
\author[5]{Portia Adade Williams}
\author[6]{Aaron Schwartz}
\author[7]{Patrick Baylis}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{Environmental Studies Program, University of Colorado Boulder}
\affil[3]{Department of Geographical Sciences, University of Maryland College Park}
\affil[4]{School of Business, East China University of Science and Technology}
\affil[5]{University of Cape Town}
\affil[6]{University of Colorado Boulder}
\affil[7]{University of British Columbia}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effects is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

\raggedbottom
\maketitle
\thispagestyle{empty}

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being and a large body of research has assessed the ongoing impacts on climate change on outcomes like agriculture, economic growth, and physical health \cite{pachauri2014climate}.  However, academic and medical researchers have highlighted the relative paucity of empirical studies on the potential impacts of climate change on mental health \cite{Berry2018Apr, hayes_climate_2018}.  While mental health is under-studied, it makes up 13\% of the global burden of disease, representing a major share of total human suffering \cite{Collins2011Jul}.  Researchers have begun to answer the call more more research, finding associations between climate disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, deteriorating environmental conditions and a sense of "ecological grief" \cite{Cunsolo2018Apr}, as well as higher temperatures and a variety of mental health impacts \cite{baylis_weather_2018, Mullins2019Dec, Li2020Mar, Obradovich2018Oct}.  Nevertheless, as global warming may reach 1.5\textdegree C within a decade \cite{allen2019technical}, more research is needed into the impacts of heat on mental health, especially in identifying the most vulnerable communities.

%Theoritical: frameworks and pathways
A number of frameworks have been proposed highlighting the pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr, BerryETAL2010}.  Work on biological mechanisms emphasizes that there are mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul, sadiq_impact_2019}.  Additionally, recent studies have found that nighttime temperatures affect sleep quality, with consequences for mental health \cite{Obradovich2017May, Mullins2019Dec}.  It has also been noted that injuries and physical health impacts from heat can have consequences for mental health \cite{Berry2007, WHO2007}.  Finally, higher temperatures affect productivity and income \cite{kjellstrom_impact_2016, Burke2015Nov}, leading to second-order impacts on mental health \cite{Katz1997, CohnETAL2004, BouchamaETAL2007}.

%Empirical: Heat and mental health
Building on these frameworks, an emerging literature has shown that higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other work demonstrating similar relationships in locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}.  Higher temperatures have also been correlated with increases in hospitalization events related to both general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb} as well as increased mortality among people with mental health disorders \cite{Hansen2008Oct}.

While the effects of heat on mental health are now well established, research conducted at national scales has found surprisingly little heterogeneity in impacts, with consequences for our understanding of vulnerability and adaptation.  For example, focusing on suicides, Burke et al. found ``no significant difference in suicide response to temperature between rich and poor municipalities or counties" \cite{Burke2018Aug}.  Similarly, Mullins et al. find no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status \cite{Mullins2019Dec}.  However, one major challenge for these studies is that they rely on data from public health agencies that is aggregated to the municipality or county level.  As a consequence, they are restricted to between-county metrics of vulnerability and cannot take neighborhood effects into account.

These findings of uniform vulnerability to heat from from national studies are surprising, because the poor and disadvantaged are much more exposed to undesirable temperatures.  They are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycle, or walk instead of commute in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality air-conditioning \cite{Samuelson2020Jun}.  Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, and these heterogeneities in vulnerability might be most pronounced within cities rather than between cities.  Indeed, neighborhoods have a large effect on vulnerability for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar}, as well as for the mental health impacts of other types of climate shocks, such as hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

While public health data on mental health outcomes is not available at the neighborhood scale, data from the social media platform Twitter can provide an important indicator of mental health at extremely precise spatial and temporal scales.  Many studies have used the mood expressed in tweets as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}.  Such studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder in individuals, even before their formal diagnosis \cite{Reece2017Oct}, while similar methods with Facebook data have been used to predict the onset of depression \cite{Eichstaedt2018Oct}.  Additionally, in London, day-to-day changes in mental health indicators derived from tweets have been associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}, while across the US, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}.  Finally, the mood expressed in tweets has been strongly associated with local weather \cite{baylis_weather_2018, hannak_tweetin_2012}, and has been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}.  Thus, Twitter provides an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures.  We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}.  Controlling for a variety of fixed effects, we find that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree C (a dry bulb temperature typically around 12\textdegree C/54\textdegree F), and the greatest declines in sentiment are between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We run models using multiple sentiment indicators, and find similar results no matter which sentiment metric was used (See Supplement).

@@

We then explore how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment (See Fig \ref{fig:hetero}).  We find that neighborhood income strongly moderates the relationship between temperature and sentiment.  As temperatures increase to a moderate heat of 20\textdegree C WBGT (29\textdegree C/84\textdegree F), sentiment decreases for most neighborhoods, but increases in neighborhoods in the 95th income percentile.  These rich areas do not see decreases in sentiment until temperatures are above 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point sentiment decreases evenly across income ranges, still leaving a large gap in expressed sentiment from low to high-income neighborhoods.

We also examine patterns of temperature and sentiment for neighborhoods that had a majority population of four broad racial categories - non-Hispanic black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, and Asian-American and Pacific Islander.  While people in neighborhoods of all racial compositions are affected by heat, people in majority black neighborhoods are much more affected than others  (See Fig \ref{fig:hetero}).  Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decrease four times as much as people in other neighborhoods.  Additionally, at levels of moderate heat from 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap closes with high temperatures.

Finally, we model the combined effects of race and income in moderating the impacts of heat on mental health, and we find separate effects for these two variables (See Supplement).  In other words, neither race nor income alone account for all the heterogeneity in vulnerability, and neighborhoods that are poor and black are more affected by heat than than neighborhoods that are just poor or just black.

\begin{figure}[H]
\centering
  \includegraphics[width=\linewidth]{../res/wbgt_combined.png}
    \caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood income (A) and race (B).}
\label{fig:hetero}
\end{figure}

@@

Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While a variety of related metrics can be used to quantify sentiment, we use the VADER metric that was specifically designed for microblogs like Twitter \cite{hutto2014vader}.  To give context to this novel metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to other events that have large impacts on sentiment (See Fig. \ref{fig:compare}).  First, we examine the change in average sentiment from Saturday to Monday, the highest and lowest days for Twitter sentiment as well as other mental health metrics, such as suicide rates \cite{CDC2021}.  We also examine the decrease in sentiment associated with the two most expensive hurricanes in the last decade in the United States.  For the hurricanes, we compare the mean sentiment of counties affected by Hurricanes Harvey and Sandy on the day of landfall to the mean sentiment exactly one week before.  These hurricanes are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

We find that the impacts of heatwaves on sentiment in rich and majority white neighborhoods are less than the average weekly change in sentiment from Saturday to Monday, but the effects of heat waves on majority black or poor neighborhoods are much greater than the average weekly change in sentiment.  Additionally, the impacts of heat waves on more vulnerable neighborhoods are close in severity to the impact of a major hurricane.

@@

To further explore our findings, we examine the impact of heat on sentiment and sentiment inequalities by Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) with over 1 million people (hereafter: cities).  We examine the impacts of heat alone, as well as the change in size of the gap in sentiment scores between both poor and rich neighborhoods and black and non-black neighborhoods.  For our metric of inequality, we show the size of the gap in sentiment as temperatures increase from optimum temperatures of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F).

Examining the effects of heat on sentiment by city, we find that heat is associated with worsened sentiment in 83.6\% of cities, and the effect is significant in 35.3\% of those.  In no cities does heat significantly increase sentiment, and heat affects sentiment in cities across the US, from Florida to New England to California.

We find that heat increases inequalities in sentiment across income groups for 68.9\% of cities, with 35.7\% of these having a statistically significant effect.  Cities with a significantly unequal effect were found in the south, Midwest, northeast, northwest, and southwest, although they were most common in the mid-Atlantic region.  Additionally, many cities in the southwest had an inverted effect, where higher temperatures actually narrowed the gap in sentiment between rich and poor neighborhoods, including Denver and Oklahoma City, which had a statistically significant inverted effect.

Because we found a much stronger effect of temperature on sentiment in majority black neighborhoods compared to other ethnic groups, we also examined increases in sentiment gaps between neighborhoods in the 95th percentile for black population (77.2\% black) and neighborhoods in the 5th percentile for black population (0\% black).  We excluded cities with a black population less than 5\% of the total population.  We found increasing inequalities in sentiment for more black neighborhoods in 64.7\% of cities, with a significant effect in 21.2\% of those (7 cities).  Additionally, three sunbelt cities had a significant inverted effect.  The patterns of inequalities in the impacts of heat on sentiment by race were similar to inequalities by income: cities with large and significant inequalities were found the mid-Atlantic and Midwest.

\begin{figure}[H]
\centering
  \includegraphics[width=0.5\linewidth]{../res/map_combined.png}
  \caption{}
  \label{fig:map}
\caption{Figure A shows the change in sentiment for a 25-degree increase in WBGT for cities with a population of over 1 million.  Figures B and C show inequalities in impacts of temperature on sentiment for cities with over 1 million people.  The value shown is the predicted change in the size of the gap in VADER score as temperatures increase from 5\textdegree C WBGT to 30\textdegree C WBGT between the 5th and 95 percentile for income (A) and the 95th and 5th percentile for percent black (B).  Cities with a statistically significant effect ($p < 0.05$) are outlined in black.  For racial inequalities (B), only cites where black people make up $> 5\%$ of the population are included.}
\end{figure}

\section*{Discussion}
% Benefits of twitter data/overview of findings
We found that twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution.  We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between heat and worsened sentiment.   Moreover, we were able to show heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

% Sentiment is a fuzzy indicator but we have BIG DATA
One limitation of data from twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of 243 million tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

% Compare weekly changes in sentiment to weekly changes in suicides
We found that the change from optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to a heatwave of 25\textdegree C WBGT (36\textdegree C/97\textdegree F) is associated with an overall decrease in the VADER sentiment score of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372), a weekly change associated with a increase in suicides of 23\% over the course of the study period \cite{CDC2021}.  Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment and suicides at some time scales.  Moreover, like sentiment, suicides are affected by higher temperatures.  Thus, while twitter sentiment is the only mental health indicator available at fine enough spatial and temporal scales to examine neighborhood effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations, although this analysis cannot speak to those outcomes specifically.
